Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-12 _tex/paper2.tex
Built site for gh-pages

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, creating a statistical phenomenon known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the existence of such hierarchical structures, the
statistical handling of this additional source of dependence between
stimuli has been inadequate. For instance, when CJ data incorporates
multiple samples of stimuli from the same individuals, researchers
frequently rely on (average) estimated BTL scores to conduct subsequent
analyses and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

@@

Researchers then typically use these estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis tests. For
example, these scores have been used to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we call the \emph{conceptual-population
model}. This model assumes that researchers have access to a conceptual
population of comparison data, that is, data representing all repeated
comparisons made by every available judge for each pair of stimuli
produced by each pair of individuals in the population, hence its name.
Conversely, Section~\ref{sec-theory-theoretical_SC} integrates the
assessment design features into what we call the \emph{sample-comparison
model}. This model assumes a more realistic scenario where researchers
only have access to a sample of judges, individuals, stimuli, and
comparisons from the conceptual population.

@@

SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
theoretical principles into the conceptual-population model. This SCM
maps the relationship between the conceptual-population outcome,
\((O^{cp}_{abhijk})\), and several related variables. The subscripts
\(a\) and \(b\) represent the compared texts (stimuli), while the
indices \(h\) and \(i\) label the students who authored them
(individuals). Moreover, the index \(j\) refers to the judge, and \(k\)
specifies the judgment number, accounting for experimental conditions
where judges compare the same texts written by the same students
multiple times, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}. It is important to note that the
indexing system allows for comparisons between different texts written
by the same student (\(h = i\), \(a \neq b\)) and between texts written
by distinct students (\(h \neq i\), where \(a = b\) is possible).
However, it excludes cases where judges compare a student's text to
itself (\(h = i\), \(a = b\)), as such a comparison lacks practical
meaning.

The other variables in SCM~\ref{fig-cj03_scm} include the texts'
discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
difference \((D_{abhijk})\), in line with Thurstone's theory (see
Section~\ref{sec-thurstone_theory}). Additionally, based on the
arguments presented in Section~\ref{sec-theory-issue1b} and the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
the SCM incorporates the judges' biases \((B_{kj})\). Together with the
outcome, these variables form the preliminary set of endogenous
variables,
\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
Finally, the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\caption{\label{fig-cj03}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

Figure~\ref{fig-cj03_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ's experiments. It also distinguishes between observable
endogenous variables, such as the outcome, and variables considered
latent according to the theory and practice of CJ, such as the text's
discriminal processes, their discriminal difference, and the judges'
biases.

Although SCMs and DAGs do not necessitate specifying the data structure
of the modeled system, defining these structures can sometimes enhance
their clarity and simplify their notation. Thus, we organize certain
indices into vectors or matrices. First, we define the matrices \(IA\)
and \(JK\) as follows:

@@

where each row of \(IA\) corresponds to a unique pairing of individuals
\(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents the number of
individuals, while \(n_{A}\) denotes the number of stimuli available per
individual. Thus, the matrix \(IA\) is composed of natural numbers
\(\mathbb{N}\), with \(n_{I} \cdot n_{A}\) rows and \(2\) columns.
Similarly, each row of \(JK\) specifies a unique pairing of a judge
\(j\) with a judgment \(k\). In this case, \(n_{J}\) indicates the total
number of judges, and \(n_{K}\) represents the number of judgments each
judge performs. Therefore, the matrix \(JK\) is also composed of natural
numbers \(\mathbb{N},\) with \(n_{J} \cdot n_{K}\) rows and \(2\)
columns.

@@

Here, the \(RPUV\) matrix is also composed of natural numbers
\(\mathbb{N}\), and has a dimension of
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
columns. The dimension of the matrix results from the
conceptual-population model's assumption that researchers have access to
data representing all repeated comparisons made by every available judge
for each pair of stimuli produced by each pair of individuals in the
population. Here, the term \({n_{I} \cdot n_{a} \choose 2}\) represents
the Binomial coefficient, which quantifies the number of
individual-stimuli comparison pairs, and it is defined as:

@@

Using the motivating example outlined in Section~\ref{sec-theoretical},
we can visualize the shape of these structures. For illustrative
purposes, assume the conceptual population consists of \(n_{I} = 5\)
individuals, each with \(n_{A} = 2\) available stimuli. Additionally,
assume that each pair of stimuli is evaluated by \(n_{J} = 3\) judges,
with each judge making \(n_{K} = 3\) comparative judgments. Given this
setup, the matrices in (\ref{eq-mat12}) and (\ref{eq-mat3}) are
represented as follows:

\[
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix}\; ; \quad
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} \; ; \quad
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

With these matrices we can express an equivalent version of
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a vectorized form.

@@

\caption{\label{fig-cj04}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

@@

\caption{\label{fig-cj09}CJ's outcome, discriminal difference,
discriminal processes and judges' biases}

@@

>>>>

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, creating a statistical phenomenon known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the existence of such hierarchical structures, the
statistical handling of this additional source of dependence between
stimuli has been inadequate. For instance, when CJ data incorporates
multiple samples of stimuli from the same individuals, researchers
frequently rely on (averaged) estimated BTL scores to conduct subsequent
analyses and tests at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

@@

Researchers then typically use these estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis tests. For
example, these scores have been used to identify `misfit' judges and
stimuli \citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
detect biases in judges' ratings
\citep{Pollitt_et_al_2003, Pollitt_2012b}, calculate correlations with
other assessment methods \citep{Goossens_et_al_2018, Bouwer_et_al_2023},
or test hypotheses related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

@@

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we call the \emph{conceptual-population
model}. This model assumes that researchers have access to comparison
data from a conceptual population, that is, data representing all
repeated judgments made by every available judge for each pair of
stimuli produced by each pair of individuals in the population, hence
its name. Conversely, Section~\ref{sec-theory-theoretical_SC} integrates
the assessment design features into what we call the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
theoretical principles into the conceptual-population model of the
motivating example. This SCM maps the relationship between the
conceptual-population outcome \((O^{cp}_{abhijk})\) and several related
variables. The subscripts \(a\) and \(b\) represent the compared texts
(stimuli), while the indices \(h\) and \(i\) label the students who
authored them (individuals). Moreover, the index \(j\) refers to the
judge, and \(k\) specifies the judgment number, accounting for
experimental conditions where judges compare the same texts written by
the same students multiple times, i.e., a \emph{repeated measures
design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system
allow comparisons between different texts written by the same student
(\(h = i\), \(a \neq b\)) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself (\(h = i\), \(a = b\)),
as such a comparison lacks practical meaning within the CJ framework.

The other variables present in SCM~\ref{fig-cj03_scm} include the texts'
discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
difference \((D_{abhijk})\), in line with Thurstone's theory (see
Section~\ref{sec-thurstone_theory}). Additionally, based on the
arguments presented in Section~\ref{sec-theory-issue1b} and the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
the conceptual-population model incorporates the judges' biases
\((B_{kj})\). Together with the outcome, these variables form the
preliminary set of endogenous variables,
\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
Finally, the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\caption{\label{fig-cj03}Conceptual-population model, preliminary
integration of theoretical principles.}

@@

Figure~\ref{fig-cj03_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. This graph illustrates the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and the practical aspects of CJ's experiments. It also
distinguishes between observable endogenous variables, such as the
outcome, and variables considered latent within the CJ framework, such
as the text's discriminal processes, their discriminal difference, and
the judges' biases.

Although specifying a system's data structure is not necessary
requirement for using SCMs and DAGs, defining one can improve clarity
and simplify their notation. Thus, we define specific data structures to
facilitate the description of the CJ system. First, we define the
matrices \(IA\) and \(JK\) as follows:

@@

where each row of \(IA\) corresponds to a unique pairing of individuals
\(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents the number of
individuals, while \(n_{A}\) denotes the number of stimuli available per
individual. Thus, the matrix \(IA\) is composed of natural numbers
\(\mathbb{N}\), with \((n_{I} \cdot n_{A})\) rows and \(2\) columns.
Similarly, each row of \(JK\) specifies a unique pairing of a judge
\(j\) with a judgment \(k\). In this case, \(n_{J}\) indicates the total
number of judges, and \(n_{K}\) represents the number of judgments each
judge performs. Therefore, the matrix \(JK\) is also composed of natural
numbers \(\mathbb{N},\) with \((n_{J} \cdot n_{K})\) rows and \(2\)
columns.

@@

The \(RPUV\) matrix is also composed of natural numbers \(\mathbb{N}\),
and has a dimension of
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
columns. The dimension of the matrix results from the
conceptual-population model's assumption that researchers have access to
data representing all repeated judgments made by every available judge
for each pair of stimuli produced by each pair of individuals in the
population. Here, the term \({n_{I} \cdot n_{A} \choose 2}\) represents
the Binomial coefficient, which quantifies the total number of
individual-stimuli comparison pairs in the conceptual population. The
coefficient is defined as:

@@

Readers can better visualize the structure of these matrices by
considering relatively small values for the parameter corresponding to
number of individuals \((n_{I} = 5)\), the number of available stimuli
per individual \((n_{A} = 2)\), the total number of judges
\((n_{J} = 3)\), and the number of judgments each judge performs
\((n_{K} = 3)\). With these values, the matrices described in equations
(\ref{eq-mat12}) and (\ref{eq-mat3}) are represented as follows:

\[
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix}\; ; \quad
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} \; ; \quad
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

Thus, with the data structure defined by the matrices in equations
(\ref{eq-mat12}) and (\ref{eq-mat3}), we can now re-express the
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
vectorized form, as shown in Figure~\ref{fig-cj04}. In this form, the
outcome \(O^{cp}_{RPUV}\), the texts' discriminal difference
\(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the judges'
biases \(B_{JK}\) are no longer scalar values but are instead vectors.
These vectors encompass all the values from the conceptual population.
Specifically, \(O^{cp}_{RPUV}\) and \(D_{RPUV}\) are vectors with a
dimension of \({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\)
values each, while \(T_{IA}\) and \(B_{JK}\) have dimensions of
\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\) values,
respectively.

@@

\caption{\label{fig-cj04}Conceptual-population model, vectorized form of
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag}.}

@@

Finally, building on the arguments developed in
Section~\ref{sec-theory-issue2} and the principles of Structural
Equation Modeling (SEM) \citep[pp.~138]{Hoyle_et_al_2023} and Item
Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model includes two (hierarchical) structural
components. Each component defines how different manifest or latent
variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition enables analyses that
simultaneously account for the traits and their associated
uncertainties, linking the trait measurement process with hypothesis
testing, thereby reducing bias and enhancing the measurement precision.

Figure~\ref{fig-cj09} displays the two (hierarchical) structural
components in the motivating example. Specifically, the
DAG~\ref{fig-cj09_dag} shows that the first component describes how the
students' overall writing quality trait \(T_{I}\) and a set of
\emph{relevant} text-related variables \(X_{IA}\) causally influence the
texts' written quality trait \(T_{IA}\). The DAG also highlights the
idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
quality that neither the students' traits nor other relevant
text-related variables can explain. In this description, \(T_{I}\) is a
vectors of dimension \(n_{I}\), \(e_{IA}\) is a vector of dimension
\((n_{I} \cdot n_{A})\), while \(X_{IA}\) is a matrix with
\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.

At a higher level in the hierarchy, the DAG~\ref{fig-cj09_dag} also
describes how a set of \emph{relevant} student-related variables
\(X_{I}\) causally influence the students' overall writing quality trait
\(T_{I}\). Additionally, the DAG accounts for idiosyncratic errors
\(e_{I}\) that capture variations in the students' writing quality that
the relevant student-related variables cannot explain. Here, \(e_{I}\)
is a vector of dimension \(n_{I}\), and \(X_{I}\) is a matrix with
\(n_{I}\) rows and \(p_{I}\) columns.

Similarly, the DAG~\ref{fig-cj09_dag} shows that the second component
describes how the judges' overall bias \(B_{J}\) and a set of
\emph{relevant} judgment-related variables \(Z_{JK}\) causally influence
the biases associated with each text \(B_{JK}\). The DAG also highlights
the idiosyncratic errors \(e_{JK}\) that capture variations in the
texts' biases that neither the judges' bias nor other relevant
judgment-related variables can explain. In this description, \(B_{J}\)
is a vectors of dimension \(n_{J}\), \(e_{JK}\) is a vector of dimension
\((n_{J} \cdot n_{K})\), while \(Z_{JK}\) is a matrix with
\((n_{J} \cdot n_{K})\) rows and \(p_{JK}\) columns.

Furthermore, at a higher level, the DAG~\ref{fig-cj09_dag} describes how
a set of \emph{relevant} judge-related variables \(Z_{J}\) causally
influence the judges' overall bias \(B_{J}\). The DAG also accounts for
idiosyncratic errors \(e_{J}\), which capture the variations in the
judges' biases that the relevant judge-related variables cannot explain.
Here, \(e_{J}\) is a vector of dimension \(n_{J}\), and \(Z_{J}\) is a
matrix with \(n_{J}\) rows and \(p_{J}\) columns.

@@

\caption{\label{fig-cj09}Conceptual-population model, full integration
of theoretical principles.}

@@

A cautious reader will notice that these descriptions rely on several
sets of relevant variables. In this study, a \emph{relevant} or a
\emph{sufficient adjustment} set of variables (potentially an empty set)
refer to those that satisfy the \emph{backdoor criterion}
\citep[pp.~37]{Neal_2020}, that is, variables that block all non-causal
paths without introducing new ones
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.

Incorporating these variables into the conceptual-population model
enables researchers to identify different estimands of interest and
their causal effects, even with observational data. This identification
works because conditioning on these variables ensures that the estimator
meets several critical properties, including confounding elimination
\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
the statistical analysis to a subset of the population defined by the
values of a given variable \citep[pp.~32]{Neal_2020}.

Naturally, the validity of causal claims now hinges on the assumption
that these variables serve as a sufficient adjustment set, a property
that cannot be guaranteed for any variable. Nevertheless, as
\citet[pp.~150]{Kohler_et_al_2019} emphasized, drawing conclusions about
the real world from observed data inevitably requires assumptions, and
this requirement holds true for both observational and experimental
data.
